\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-27 chapters/abstract_fr.tex
Typos françois

Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les
techniques générales de décodage ERP en affinant la structure des estimateurs
linéaires et multilinéaires.
Ces méthodes améliorent la précision de classification dans une gamme de
conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et
multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et contribue à la generalisation, ce qui
permets le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
contextes et besoins des utilisateurs.

>>>>

Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a pallier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de l'attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

Au-delà de l'indépendance du regard, le travail fait également progresser les
techniques générales de décodage ERP en affinant la structure des estimateurs
linéaires et multilinéaires.
Ces méthodes améliorent la précision de classification dans une gamme de
conditions BCI, en particulier lorsque les données d'entraînement sont limitées.
L'introduction de régularisation structurée dans les modèles linéaires et
multilinéaires améliore l'interprétabilité des classificateurs et réduit le temps d'entraînement ainsi que la complexité computationnelle.
Cela permet un entraînement plus efficace et contribue à la g\'en\'eralisation, ce qui
permet le d\'evelopment des systèmes plus fiables qui s'adaptent à divers
contextes et besoins des utilisateurs.

### 2024-11-27 chapters/bci.tex
Layout chapter gaze independence
Typos françois

\Acp{bci} Started as a speculative concept in the 1970's~\cite{Vidal1973}.
Now, they have gradually evolved from methods  developed in the the 1990's to
establish basic communication from minimal brain activity
(yes/no, left/right, \ldots)~\cite{Wolpaw2002} into sophisticated interaction
schemes that restore or enhance many capabilities by bypassing defective parts
of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}
~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and spatial resolution.
  }
  \label{fig:bci/recording}
}

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can be
calculated as
\begin{equation}
  \text{\acs{itr}} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of different options
that can be selected per trial\footnote{Given that selections are independent of each
other. This formula requires some adaptations in the case of sequential or hierarchical
selection.}, $P$, the selection accuracy of the decoder, and $Q$, the number of trials
per second.
The parameters $N$, $P$, and $Q$ of this formula give us some insight into the building
blocks of a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a broad range of
informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for decoding,
and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or responses.
\end{enumerate*}

@@

Linear decoding techniques such as the \ac{lcmv} beamformer~\cite{Wittevrongel2016}
generally work well for classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}\in\mathbb{R}^{CS}$ from
a set of epochs that can be multiplied with the feature vector $\mat{x}$ of a
new epoch to obtain a prediction score.
The weights are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with \emph{activation pattern} $\mat{a}\in\mathbb{R}^{CS}$ the difference between the
average of feature vectores of the attended and non-attended epochs in the
training data,
and $\mat{C}\in\mathbb{R}^{CS\times CS}$ the background covariance matrix which
can be estimated empirically from the training data.
This is then solved by the method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}

>>>>

\Acp{bci} started as a speculative concept in the 1970's~\cite{Vidal1973}.
Now, they have gradually evolved from methods  developed in the the 1990's to
establish basic communication from minimal brain activity
(yes/no, left/right, \ldots)~\cite{Wolpaw2002} into sophisticated interaction
schemes that restore or enhance many capabilities by bypassing defective parts
of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}
~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and \\ spatial resolution.
  }
  \label{fig:bci/recording}
}

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can be
calculated as
\begin{equation}
  \ITR = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of different options
that can be selected per trial\footnote{Given that selections are independent of each
other. This formula requires some adaptations in the case of sequential or hierarchical
selection.}, $P$, the selection accuracy of the decoder, and $Q$, the number of trials
per second.
The parameters $N$, $P$, and $Q$ of this formula give us some insight into the building
blocks of a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a broad range of
informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for decoding,
and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or responses.
\end{enumerate*}

@@

Linear decoding techniques such as the \ac{lcmv} beamformer~\cite{Wittevrongel2016}
generally work well for classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}\in\mathbb{R}^{CS}$ from
a set of epochs that can be multiplied with the feature vector $\mat{x}$ of a
new epoch to obtain a prediction score.
The weights are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with \emph{activation pattern} $\mat{a}\in\mathbb{R}^{CS}$ the difference between the
average of feature vectors of the attended and non-attended epochs in the
training data,
and $\mat{C}\in\mathbb{R}^{CS\times CS}$ the background covariance matrix which
can be estimated empirically from the training data.
This is then solved by the method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}

### 2024-11-27 chapters/conclusion.tex
Typos françois

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
As the work presented in this thesis started during the COVID-19 pandemic,
experimentation was delayed and the envisioned application could not be
validated within the time frame of the doctoral project.
To complete this work, we aim to revisit the participants with an on-line
experiment and satisfaction assessment to establish the usability of the proposed
interface according the principles of \ac{ucd} and
compare its \ac{itr} to the state-of-the-art literature.

>>>>

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
As the work presented in this thesis started during the COVID-19 pandemic,
experimentation was delayed and the envisioned application could not be
validated within the time frame of the doctoral project.
To complete this work, we aim to revisit the participants with an on-line
experiment and satisfaction assessment to establish the usability of the proposed
interface according to the principles of \ac{ucd} and
compare its \ac{itr} to the state-of-the-art literature.

### 2024-11-27 chapters/gaze_independence.tex
Typos françois

\chapter{Gaze-independent visual \acsp{bci}}
\label{sec:gaze-independent}
\epigraph{%
  \justifying
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

Another other group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
Additionally, individuals with other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

@@

To test this hypothesis, we chose a suitable interface, innovate on decoder
development and test if these innovations improve \ac{bci} performance in a
meaningful way.
Accordingly, we collected data from healthy control subjects to test proposed
decoders, and the findings have be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis}, focusing
both accuracy and comfort.

>>>>

\chapter{Gaze-independent visual \acsp{bci}}
\label{sec:gaze-independent}
\epigraph{%
  %\justifying
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

Another group that is heavily affected, is \ac{cp}~\cite{Fazzi2012}.
Additionally, individuals with other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

@@

To test this hypothesis, we chose a suitable interface, innovate on decoder
development and test if these innovations improve \ac{bci} performance in a
meaningful way.
Accordingly, we collected data from healthy control subjects to test proposed
decoders, and the findings have been verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis}, focusing
both accuracy and comfort.

### 2024-11-27 cover_ulille.tex
Typos françois

      Co-directeur de th\`ese: \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
      Co-directeur de th\`ese: \emph{Prof.\ Marc Van Hulle, KU Leuven, D\'epartment des Neurosciences} \\
      Pr\'esident: \emph{Prof.\ Koen Poesen, KU Leuven, D\'epartment des Neurosciences} \\
      Rapporteuse: \emph{Prof.\ Andrea K\"ubler, Univ. W\"urzburg, Institut de Psychologie} \\
      Rapporteur: \emph{Prof.\ Fabien Lotte, Inria Bordeaux, LaBRI} \\
      Rapporteur: \emph{Prof.\ Reinhold Scherer, Univ. Essex, École d'informatique et de G\'enie \'Electronique} \\
      Examinateur: \emph{Prof.\ Maarten De Vos, KU Leuven,  D\'epartment de G\'enie \'Electrique} \\
      Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique}
}

>>>>

     Pr\'esident: \emph{Prof.\ Koen Poesen, KU Leuven, D\'epartment des Neurosciences} \\
      Rapporteuse: \emph{Prof.\ Andrea K\"ubler, Univ. W\"urzburg, Institut de Psychologie} \\
      Rapporteur: \emph{Prof.\ Fabien Lotte, Inria Bordeaux, LaBRI} \\
      Rapporteur: \emph{Prof.\ Reinhold Scherer, Univ. Essex, École d'informatique et de G\'enie \'Electronique} \\
      Examinateur: \emph{Prof.\ Maarten De Vos, KU Leuven,  D\'epartment de G\'enie \'Electrique} \\
      Examinateur: \emph{Prof.\ Adalberto Simeone, KU Leuven, D\'epartment d'Informatique} \\
      Co-directeur de th\`ese (invit\'e): \emph{Prof.\ François Cabestaing, Univ. Lille, UMR CRIStAL} \\
      Co-directeur de th\`ese (invit\'e): \emph{Prof.\ Marc Van Hulle, KU Leuven, D\'epartment des Neurosciences}

}
